---
title: "Progress Memo"
subtitle: |
  | Final Project 
  | Data Visualization (STAT 302)
author: "Saul Pink"
date: today

format:
  html:
    toc: true
    embed-resources: true
    
execute:
  echo: false
  warning: false
---

```{r}
#| label: load-pckgs

# load packages
library(tidyverse)
library(here)
library(skimr)
library(knitr)

```

## Data source

Response here.

## Basic data description

```{r}
#| label: tbl-read-data
#| tbl-cap: Sample of first five observations from NCHS Leading Causes of Death database
#| tbl-cap-location: bottom

# read in data
death_causes <- read_csv(here("data/leading_causes_of_death.csv")) |> 
  janitor::clean_names()

# build table
death_causes |> 
  slice_head(n = 5) |> 
  kable(align = "c")

```

@tbl-read-data shows the first five observations from the dataset to demonstrate that it was succesfully read in. We see that there are six columns. The key columns we are interested in visualizing are `year`, `cause_name`, `state`, and `age_adjusted_death_rate`. 

```{r}
#| label: tbl-skim-data
#| tbl-cap: Skim of NCHS leading causes of death data
#| tbl-cap-location: top

# skim without charts
death_causes |> 
  skim_without_charts()

```

In @tbl-skim-data, we get an overview of the dataset. There are 10,868 observations with six variables, as noted above. There are three character variables: `x113_cause_name`, `cause_name`, and `state`. There are three numeric variables: `year`, `deaths`, and `age_adjusted_death_rate`. There is no missingness in the dataset. I will also use spatial data from the `maps` package to map the leading causes of death. 

## Why this data

Response here.

## Visualization plan

Response here.

## Misc.

Response here.

